Reproduction frequency and offspring survival decline in elephant seals past prime age

### 2024-07-15 paper/paperv3.qmd
selective (dis)appearance, bio vs chrono age

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = (age - age_senesce) / 10,
         age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old"))) %>% 
  group_by(animalID) %>% 
  mutate(firstrepro = ifelse(max(year) < 2020 && any(observed_int == 1),
                             min(age[observed_int == 1]),
                             NA),
         lastrepro = ifelse(max(year) < 2020 && any(observed_int == 1),
                            max(age[observed_int == 1]),
                            NA)) %>% 
  ungroup()

@@

Age-dependent life history shifts can result from within- or between-animal changes [@vandepol2006]. We hypothesized age-dependent changes in reproduction were due to senescence, i.e. within-animal changes. However, it is possible that selective appearance or disappearance of animals in the population can create the same pattern. For example, if animals the breed more frequently have lower survival, then older animals will reproduce less frequently not because of senescence but rather because infrequent breeders disproportionately survive to older ages (i.e. selective disappearance). We tested whether our significant results (reproductive senescence, maternal effect senescence in pup survival) were influenced by selective appearance and disappearance, following the methods of @vandepol2006, by adding terms for age at first and last reproduction, respectively. We used AIC to compare the base model to models with selective appearance and disappearance.

```{r}
#| label: selective-h1

@@

```{r}
#| label: fig-selective-h1
#| fig-cap: Predicted breeding probabilities of (**A**) selective appearance and (**B**) selective disappearance models. The semi-transparent solid lines are predictions of the base model. The dashed and dotted lines represent predictions for the 10th and 90th percentiles of observed age at first (**A**) and last (**B**) reproduction. In **A**, these correspond to first reproduction at 4 and 7 years old, respectively. In **B**, these correspond to last reproduction at 5 and 17 years old, respectively. Confidence intervals not shown for clarity.
#| fig-width: 9

# Compare model predictions
base_pred <- ggpredict(
  h1_selective_models[[1]],
  terms = c("age10 [all]", "age_cat")
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

# selective appearance for firstrepro=4,6 (10% and 90% of observed)
selapp_pred <- ggpredict(
  h1_selective_models[[2]],
  terms = c("age10 [all]", "age_cat", "firstrepro [4,7]")
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old")),
         firstrepro = facet) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

# selective disappearance for lastrepro=5,17 (10% and 90% of observed) 
seldis_pred <- ggpredict(
  h1_selective_models[[4]],
  terms = c("age10 [all]", "age_cat", "lastrepro [5,17]")
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old")),
         lastrepro = facet) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

# Selective appearance predictions
fig_selapp <- ggplot(base_pred, aes(age, predicted)) +
  # Base predictions
  geom_line(aes(color = age_cat), alpha = 0.5) +
  # Selective appearance ages 4 and 6
  geom_line(aes(color = age_cat, 
                linetype = firstrepro),
            selapp_pred) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Breeding", 
                     labels = scales::percent) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_linetype_manual(values = c("dashed", "dotted")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

fig_seldis <- ggplot(base_pred, aes(age, predicted)) +
  # Base predictions
  geom_line(aes(color = age_cat), alpha = 0.5) +
  # Selective disappearance ages 5 and 17
  geom_line(aes(color = age_cat, 
                linetype = lastrepro),
            seldis_pred) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Breeding", 
                     labels = scales::percent) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_linetype_manual(values = c("dashed", "dotted")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

(fig_selapp | fig_seldis) +
  plot_annotation(tag_levels = "A")
```

There was no evidence for selective appearance or disappearance influencing the relationship between maternal age and probability of reproduction. The selective appearance model (i.e., included a term for age-of-first-reproduction) had a similar AIC weight as the base model. However, in the selective appearance model, the coefficient for age at first reproduction was not significantly different than 0 (`r h1_firstrepro_95ci`). As a result, the predicted breeding probabilities for animals that first recruited at ages 4 and 7 (the 10th and 90th observed percentiles in the population) were not qualitatively different than the predictions of the base model [@fig-selective-h1]. Therefore the observed pattern is consistent with the senescence hypothesis (i.e., within-animal change with age).

```{r}
#| label: selective-h2

h2_aic <- do.call(AIC, h2_selective_models) %>% 
  as_tibble() %>% 
  mutate(model = c("Base", 
                   "Selective appearance", 
                   "Selective disappearance", 
                   "Selective appearance and disappearance"),
         dAIC = AIC - min(AIC),
         rellik = exp(-0.5 * dAIC),
         weight = rellik / sum(rellik)) %>% 
  relocate(model)

# Selective appearance model, first repro 95% ci
h2_selapp_coef <- summary(h2_selective_models[[2]])$coef["firstrepro", ]
h2_firstrepro_95ci <- (h2_selapp_coef["Estimate"] + c(-1.96, 1.96) * h2_selapp_coef["Std. Error"]) %>% 
  make_95()

h2_aic

```{r}
#| label: fig-selective-h2
#| fig-cap: Predicted breeding probabilities of (**A**) selective appearance and (**B**) selective disappearance models. The semi-transparent solid lines are predictions of the base model. The dashed and dotted lines represent predictions for the 10th and 90th percentiles of observed age at first (**A**) and last (**B**) reproduction. In **A**, these correspond to first reproduction at 4 and 7 years old, respectively. In **B**, these correspond to last reproduction at 5 and 17 years old, respectively. Confidence intervals not shown for clarity.
#| fig-width: 9

# Compare model predictions
base_pred <- ggpredict(
  h2_selective_models[[1]],
  terms = c("age10 [all]", "age_cat")
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

# selective appearance for firstrepro=4,6 (10% and 90% of observed)
selapp_pred <- ggpredict(
  h2_selective_models[[2]],
  terms = c("age10 [all]", "age_cat", "firstrepro [4,7]")
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old")),
         firstrepro = facet) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

# selective disappearance for lastrepro=5,17 (10% and 90% of observed) 
seldis_pred <- ggpredict(
  h2_selective_models[[4]],
  terms = c("age10 [all]", "age_cat", "lastrepro [5,17]")
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = factor(group, levels = c("Young", "Old")),
         lastrepro = facet) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce),
         age <= 20)

# Selective appearance predictions
fig_selapp <- ggplot(base_pred, aes(age, predicted)) +
  # Base predictions
  geom_line(aes(color = age_cat), alpha = 0.5) +
  # Selective appearance ages 4 and 6
  geom_line(aes(color = age_cat, 
                linetype = firstrepro),
            selapp_pred) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Offspring survival", 
                     labels = scales::percent) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_linetype_manual(values = c("dashed", "dotted")) +
  labs(x = "Maternal age (years)") +
  theme(legend.position = "none")

fig_seldis <- ggplot(base_pred, aes(age, predicted)) +
  # Base predictions
  geom_line(aes(color = age_cat), alpha = 0.5) +
  # Selective disappearance ages 5 and 17
  geom_line(aes(color = age_cat, 
                linetype = lastrepro),
            seldis_pred) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Offspring survival", 
                     labels = scales::percent) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_linetype_manual(values = c("dashed", "dotted")) +
  labs(x = "Maternal age (years)") +
  theme(legend.position = "none")

fig_selapp | fig_seldis +
  plot_annotation(tag_levels = "A")
```

There was no evidence for selective appearance or disappearance influencing the relationship between maternal age and offspring survival. The selective appearance model (i.e., included a term for age-of-first-reproduction) had a slightly greater AIC weight than the base model. However, in the selective appearance model, the coefficient for age at first reproduction was not significantly different than 0 (`r h2_firstrepro_95ci`). As a result, the predicted breeding probabilities for animals that first recruited at ages 4 and 7 (the 10th and 90th observed percentiles in the population) were not qualitatively different than the predictions of the base model [@fig-selective-h2]. Therefore the observed pattern is consistent with the senescence hypothesis (i.e., within-animal change with age).

@@

bioage_senesce <- -7
sealdat2 <- sealdat %>% 
  drop_na(lastrepro) %>% 
  mutate(bioage = age - lastrepro,
         bioage10 = (bioage - bioage_senesce) / 10,
         bioage_cat = factor(bioage < bioage_senesce,
                             labels = c("Old", "Young")))

@@

The chronological age of an individual may not reflect their biological age because aging rates vary within populations [@martin2011]. Years-to-death, a proxy for biological age, may therefore be a better predictor for senescence patterns than chronological age. We repeated our analysis for the reproductive senescence hypothesis using biological age, setting the breakpoint for senescence at biological age -7 years (i.e. 7 years prior to final observation), based on visual inspection of the raw data. This required us to limit our sample to animals with known longevity (i.e., last observation in 2020 or earlier), which reduced our sample size to `r nrow(sealdat2)` observations of `r n_distinct(sealdat2$animalID)` distinct animals. Subsequently, we fit the model for breakpoints at all biological ages -10 to -3 years, to assess whether our results were sensitive to the choice of breakpoint. Finally, we re-fit the chronological reproductive senescence model using the known-longevity dataset and compared chronological versus biological age models using AIC.

@@

breed_bioage_mod <- glmer(
  observed_int ~ bioage10 : bioage_cat + (1 | animalID) + (1 | year_fct),
  sealdat2,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

bioage_pval <- sprintf("%0.3f", summary(breed_bioage_mod)$coef["bioage10:bioage_catOld", "Pr(>|z|)"])

breed_age_mod2 <- glmer(
  observed_int ~ age10 : age_cat + (1 | animalID) + (1 | year_fct),
  drop_na(sealdat, lastrepro),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

p_by_tau <- map_dbl(-10:-3, \(tau) {
  mod <- sealdat %>%
    drop_na(lastrepro) %>%
    mutate(bioage = age - lastrepro,
           bioage10 = (bioage - tau) / 10,
           bioage_cat = factor(bioage < tau,
                               labels = c("Old", "Young"))) %>% 
    glmer(
      observed_int ~ bioage10 : bioage_cat + (1 | animalID) + (1 | year_fct),
      .,
      family = "binomial",
      control = glmerControl(optimizer = "bobyqa")
    )
  summary(mod)$coef["bioage10:bioage_catOld", "Pr(>|z|)"]
})

@@

# Summarize the raw data
bioage_summ <- sealdat2 %>% 
  group_by(bioage) %>% 
  summarize(n_obs = n(), 
            perc_breed = mean(observed_int), 
            se = sqrt(perc_breed * (1 - perc_breed) / n_obs))

# Create POPULATION-LEVEL (i.e. without rand effects) predictions
bioage_pred_pop <- ggpredict(breed_bioage_mod, 
                             terms = c("bioage10 [all]", "bioage_cat")) %>% 
  as_tibble() %>% 
  mutate(bioage = x * 10 + bioage_senesce,
         bioage_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((bioage_cat == "Young" & bioage < bioage_senesce) |
           (bioage_cat == "Old" & bioage >= bioage_senesce))

# Create WEIGHTED-AVERAGE predictions using years' random effects
bioage_pred_wgt <- ggpredict(
  breed_bioage_mod,
  terms = c("bioage10 [all]", "bioage_cat", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(bioage = x * 10 + bioage_senesce,
         bioage_cat = factor(group, levels = c("Young", "Old")),
         year_fct = facet) %>% 
  filter((bioage_cat == "Young" & bioage < bioage_senesce) |
           (bioage_cat == "Old" & bioage >= bioage_senesce)) %>% 
  left_join(seal_years, by = "year_fct") %>% 
  group_by(bioage, bioage_cat) %>% 
  summarize(predicted = weighted.mean(predicted, n),
            .groups = "drop")

ggplot(bioage_pred_pop, aes(bioage, predicted)) +
  geom_ribbon(aes(fill = bioage_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(aes(color = bioage_cat), linewidth = 1.2, linetype = "dashed") +
  geom_line(aes(color = bioage_cat),
            bioage_pred_wgt, 
            linewidth = 1.2,
            linetype = "dotted") +
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  bioage_summ) +
  geom_vline(xintercept = bioage_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Breeding probability", labels = scales::percent, limits = c(.7, 1)) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_x_continuous(limits = c(-17, 0)) + 
  labs(x = "Biological age (years-to-death)") +
  theme(legend.position = "none")

@@

The biological age model did not support the reproductive senescence hypothesis. The coefficient for biological age among older animals (biological age \> -7 years) was `r bioage_coef` (95% CI) which was not significantly different than 0 (*p* = `r bioage_pval`). Choice of breakpoint for senescence did not affect the significance of this coefficient. For breakpoints -10 to -3 years, *p* reached a minimum at biological age -7 years.

@@

@@

@@

missing_years <- count(missing_seals, year_fct)
missing_pred <- expand_grid(age = 5:19, 
                            year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10) %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = missing_grid,
                             re.form = ~ (1 | year_fct)),
         se = predict(missing_mod, 
                      newdata = missing_grid,
                      se.fit = TRUE,
                      re.form = NA)$se.fit,
         conf.low = predicted - 1.96 * se,
         conf.high = predicted + 1.96 * se) %>% 
  left_join(missing_years, by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(across(c(predicted, conf.low, conf.high), mean)) %>% 
  mutate(across(c(predicted, conf.low, conf.high), family(missing_mod)$linkinv))

missing_pred_years <- expand_grid(age = 5:19, 
                            year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10) %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = missing_grid,
                             type = "response",
                             re.form = ~ (1 | year_fct)))

missing_pred

missing_seals %>% 
  group_by(age) %>% 
  summarize(unobserved_mean = mean(unobserved),
            unobserved_se = sqrt(unobserved_mean * (1 - unobserved_mean) / n()),
            unobserved_lwr = unobserved_mean - unobserved_se,
            unobserved_upr = unobserved_mean + unobserved_se) %>% 
  ggplot(aes(age)) +
  geom_line(aes(y = predicted, group = year_fct), 
            missing_pred_years, 
            alpha = 0.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              missing_pred,
              alpha = 0.2,
              fill = "#de77ae") +
  geom_line(aes(y = predicted),
            missing_pred,
            color = "#de77ae",
            linewidth = 2) +
  geom_pointrange(aes(y = unobserved_mean, 
                      ymin = unobserved_lwr, ymax = unobserved_upr)) +
  scale_x_continuous("Age", breaks = seq(5, 19, by = 4)) +
  scale_y_continuous("Unobserved proportion", 
                     labels = scales::label_percent(accuracy = 1),
                     limits = c(0, NA))
```

>>>>

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = (age - age_senesce) / 10,
         age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old"))) %>% 
  group_by(animalID) %>% 
  mutate(firstrepro = ifelse(max(year) < 2020 && any(observed_int == 1),
                             min(age[observed_int == 1]),
                             NA),
         lastrepro = ifelse(max(year) < 2020 && any(observed_int == 1),
                            max(age[observed_int == 1]),
                            NA),
         lastobs = ifelse(max(year) < 2020, max(age), NA)) %>% 
  ungroup()

@@

Age-dependent life history shifts can result from within- or between-animal changes [@vandepol2006]. We hypothesized age-dependent changes in reproduction were due to senescence, i.e. within-animal changes. However, it is possible that selective appearance or disappearance of animals in the population can create the same pattern. For example, if animals the breed more frequently have lower survival, then older animals will reproduce less frequently not because of senescence but rather because infrequent breeders disproportionately survive to older ages (i.e. selective disappearance). We tested whether our significant results (reproductive senescence, maternal effect senescence in pup survival) were influenced by selective appearance and disappearance, following the methods of @vandepol2006, by adding terms for age at first and last reproduction, respectively. If our results are due to selective appearance/disappearance instead of senescence, then the 95% CI for the coefficient of age in the old age class in these models will overlap 0.

```{r}
#| label: fig-selective
#| fig-cap: The coefficient for age in the old age class in models accounting for selective appearance and disappearance. Red and blue indicate  reproductive senescence (H1) and maternal effect senescence (H2a), respectively. Compared to the base model, the appearance model includes a term for age at first reproduction, the disappearance model includes a term for age at last reproduction, and the combined model includes both.

@@

There was no evidence for selective appearance or disappearance influencing the relationship between maternal age and probability of reproduction or offspring survival. The 95% CI of the coefficient for age in the old age class did not overlap 0 for either hypothesis in any of the selective (dis)appearance models [@fig-selective]. Therefore the observed patterns are consistent with the senescence hypothesis (i.e., within-animal change with age).

@@

bioage_senesce <- -7
sealdat2 <- sealdat %>% 
  drop_na(lastrepro) %>% 
  mutate(bioage = age - lastobs,
         bioage10 = (bioage - bioage_senesce) / 10,
         bioage_cat = factor(bioage < bioage_senesce,
                             labels = c("Old", "Young")))

@@

The chronological age of an individual may not reflect their biological age because aging rates vary within populations [@martin2011]. Years-to-death, a proxy for biological age, may therefore be a better predictor for senescence patterns than chronological age. We repeated our analysis for the reproductive senescence hypothesis using biological age, setting the breakpoint for senescence at biological age -7 years (i.e. 7 years prior to final observation), based on visual inspection of the raw data. This required us to limit our sample to animals with known longevity (i.e., last observation in 2020 or earlier), which reduced our sample size to `r nrow(sealdat2)` observations of `r n_distinct(sealdat2$animalID)` distinct animals. Subsequently, we fit the model for breakpoints at all biological ages -12 to -3 years, to assess whether our results were sensitive to the choice of breakpoint. Finally, we re-fit the chronological reproductive senescence model using the known-longevity dataset and compared chronological versus biological age models using AIC.

@@

# Biological age model
breed_bioage_mod <- glmer(
  observed_int ~ bioage10 : bioage_cat + (1 | animalID) + (1 | year_fct),
  sealdat2,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

old_z <- summary(breed_bioage_mod)$coefficients["bioage10:bioage_catOld", "z value"]
bioage_pval <- sprintf("%0.3f", pnorm(-old_z))

# Breakpoint comparison
p_by_tau <- map_dbl(-12:-3, \(tau) {
  mod <- sealdat %>%
    drop_na(lastrepro) %>%
    mutate(bioage = age - lastobs,
           bioage10 = (bioage - tau) / 10,
           bioage_cat = factor(bioage < tau,
                               labels = c("Old", "Young"))) %>% 
    glmer(
      observed_int ~ bioage10 : bioage_cat + (1 | animalID) + (1 | year_fct),
      .,
      family = "binomial",
      control = glmerControl(optimizer = "bobyqa")
    )
  old_z <- summary(mod)$coef["bioage10:bioage_catOld", "z value"] 
  pnorm(old_z)
})

tibble(`Biological age threshold` = -12:-3,
       `p value` = p_by_tau)

# Model comparison
breed_age_mod2 <- glmer(
  observed_int ~ age10 : age_cat + (1 | animalID) + (1 | year_fct),
  drop_na(sealdat2, lastrepro),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

# Summarize the raw data
bioage_summ <- sealdat2 %>% 
  group_by(bioage) %>% 
  summarize(n_obs = n(), 
            n_breed = sum(observed_int),
            perc_breed = n_breed / n_obs, 
            breed_lwr = binom.test(n_breed, n_obs)$conf.int[1], 
            breed_upr = binom.test(n_breed, n_obs)$conf.int[2])

# Create ANNUAL (i.e. rand effect of year only) predictions
bioage_pred_years <- ggpredict(
  breed_bioage_mod, 
  terms = c("bioage10 [all]", "bioage_cat", "year_fct"),
  type = "random"
) %>% 
  as_tibble() %>% 
  mutate(bioage = x * 10 + bioage_senesce,
         bioage_cat = factor(group, levels = c("Young", "Old")),
         year_fct = factor(facet)) %>% 
  filter((bioage_cat == "Young" & bioage < bioage_senesce) |
           (bioage_cat == "Old" & bioage >= bioage_senesce),
         bioage <= 20)

# Create WEIGHTED-AVERAGE predictions using years' random effects
# ggpredict can't do that for confidence intervals
# create prediction grid
bioage_grid <- expand_grid(
  bioage = -17:0,
  year_fct = factor(1988:2019)
) %>% 
  mutate(bioage10 = (bioage - bioage_senesce) / 10,
         bioage_cat = factor(ifelse(bioage < bioage_senesce, "Young", "Old"),
                          levels = c("Young", "Old")))

bioage_years <- count(sealdat2, year_fct)
bioage_pred_wgt <- bioage_grid %>% 
  mutate(predicted = predict(breed_bioage_mod, 
                             newdata = bioage_grid,
                             re.form = ~ (1 | year_fct)),
         predicted_se = suppressWarnings(
           # warning: "se.fit computation uses an approximation to estimate the sampling distribution of the parameters"
           predict(breed_bioage_mod, 
                   newdata = bioage_grid,
                   re.form = NA,
                   se.fit = TRUE)
         )$se.fit,
         predicted_pop = predict(breed_bioage_mod, 
                                 newdata = bioage_grid,
                                 re.form = NA),
         conf_lo = predicted - 1.96 * predicted_se,
         conf_hi = predicted + 1.96 * predicted_se) %>% 
  left_join(bioage_years, by = "year_fct") %>% 
  group_by(bioage, bioage_cat) %>% 
  summarize(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                   \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  # invert logit
  mutate(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                family(breed_bioage_mod)$linkinv)) %>% 
  left_join(count(sealdat2, bioage), by = "bioage")

ggplot(bioage_pred_wgt, aes(bioage, predicted)) +
  # Predictions for individual years
  geom_line(aes(group = interaction(bioage_cat, year_fct)),
            bioage_pred_years,
            alpha = 0.1) +
  # Weighted model (CI and mean response)
  geom_ribbon(aes(fill = bioage_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.5) +
  geom_line(aes(color = bioage_cat),
            linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = bioage_cat),
            linetype = "dotted",
            linewidth = 1.2,
            alpha = 0.5) +
  # Summarized raw data
  geom_pointrange(aes(y = perc_breed,
                      ymin = breed_lwr,
                      ymax = breed_upr),
                  bioage_summ) +
  geom_vline(xintercept = bioage_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Breeding probability", 
                     labels = scales::percent,
                     limits = c(NA, 1.02),
                     breaks = seq(0.4, 1.0, by = 0.2)) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  labs(x = "Biological age (years)") +
  theme(legend.position = "none") + 
  geom_text(aes(label = n), y = 1, vjust = -0.5)

@@

The biological age model did not support the reproductive senescence hypothesis. The coefficient for biological age among older animals (biological age \> -7 years) was `r bioage_coef` (95% CI) which was not significantly different than 0 (*p* = `r bioage_pval`). Choice of breakpoint for senescence did not affect the significance of this coefficient. For breakpoints -12 to -3 years, *p* reached a minimum at biological age -7 years (see table above).

@@

::: callout-note
Max's commentary: almost all seals bred in the last year they were observed. As a result, 96.6% (95% CI: 95.2% - 97.7%) of seals in the bioage 0 group bred. That proportion was greater than all bioages except -17 and -16, which had much smaller sample sizes (0: 903, -17: 1, -16: 4). With such a high probability in bioage 0, it's just about impossible to get a negative slope in the old age category, because that one data point is pulling the right side of the line up. FWIW, 53.7% of the bioage 0 animals have a chronological age 4-7, so that part of the "old" biological age category is mostly chronologically young animals.
:::

@@

missing_age_se <- summary(missing_mod)$coef["age10", "Std. Error"]
missing_age_est <- summary(missing_mod)$coef["age10", "Estimate"]
missing_age_coef <- sprintf("%0.2f (95%% CI: %0.2f - %0.2f)",
                            missing_age_est,
                            missing_age_est - 1.96 * missing_age_se,
                            missing_age_est + 1.96 * missing_age_se)

@@

missing_years <- count(missing_seals, year_fct)
missing_grid <- expand_grid(age = 5:19, 
                            year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10)

missing_pred <- missing_grid %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             re.form = ~ (1 | year_fct)),
         se = suppressWarnings(predict(missing_mod, 
                                       newdata = .,
                                       se.fit = TRUE,
                                       re.form = NA)$se.fit),
         conf.low = predicted - 1.96 * se,
         conf.high = predicted + 1.96 * se) %>% 
  left_join(missing_years, by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(across(c(predicted, conf.low, conf.high), mean)) %>% 
  mutate(across(c(predicted, conf.low, conf.high), family(missing_mod)$linkinv))

missing_pred_years <- missing_grid %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             type = "response",
                             re.form = ~ (1 | year_fct)))

missing_seals %>% 
  group_by(age) %>% 
  summarize(n = n(),
            n_unobserved = sum(unobserved),
            unobserved_mean = n_unobserved / n,
            unobserved_lwr = binom.test(n_unobserved, n)$conf.int[1],
            unobserved_upr = binom.test(n_unobserved, n)$conf.int[2]) %>% 
  ggplot(aes(age)) +
  geom_line(aes(y = predicted, group = year_fct), 
            missing_pred_years, 
            alpha = 0.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              missing_pred,
              alpha = 0.2,
              fill = "#de77ae") +
  geom_line(aes(y = predicted),
            missing_pred,
            color = "#de77ae",
            linewidth = 2) +
  geom_pointrange(aes(y = unobserved_mean, 
                      ymin = unobserved_lwr, ymax = unobserved_upr)) +
  scale_x_continuous("Age", breaks = seq(5, 19, by = 4)) +
  scale_y_continuous("Unobserved proportion", 
                     labels = scales::label_percent(accuracy = 1),
                     limits = c(0, NA))
```
